
% --- Title block ---
\begin{center}
  {\LARGE\bfseries rvLLM: A High-Performance LLM Inference Engine in Rust\par}
  \vspace{0.6em}
  {\large Andy Norris (m0at)\par}
  \vspace{0.3em}
  {\normalsize\texttt{https://github.com/m0at/rvllm}\par}
  \vspace{0.3em}
  {\normalsize San Francisco, April 14, 2026\par}
\end{center}
\vspace{1em}

% --- Abstract ---
\begin{abstract}
I present rvLLM, a from-scratch Rust implementation of the vLLM inference engine
for OpenAI-compatible API serving with CUDA GPU acceleration. The current codebase
is a 31-crate Rust workspace ($\sim$76K lines) with 54 CUDA kernels, 13 model
architectures (Llama, Qwen2, Mistral, DeepSeek-V2, Mixtral, NemotronH~MoE, Gemma,
and others), FA3 v3 cp.async + split-KV attention, no-fallback kernel validation,
and a Rust-native PTX compiler that generates fused decode kernels
2--7.5$\times$ faster than the hand-written CUDA baselines in the measured
microbenchmarks. In the latest same-GPU comparison on H100~SXM~80\,GB
serving Qwen2.5-7B against vLLM~0.19.0 FP8 (April 2026), rvLLM~v2 FP8 reaches
29{,}868~tok/s at $N\!=\!64$ (\textbf{2.2$\times$ vLLM}),
30{,}076~tok/s at $N\!=\!128$ (\textbf{1.3$\times$ vLLM}), and
13{,}865~tok/s at $N\!=\!1$ (\textbf{79$\times$ vLLM}).
The key changes in this revision are: CUTLASS~3x SM90 FP8 GEMMs with Hopper WGMMA
and TMA replacing cuBLASLt, fused RMSNorm+FP8 quantization kernels, fused
SiLU+FP8 quantization kernels, vectorized fused SiLU with 128-bit loads and
64-bit FP8 stores, and stream-K/split-K CUTLASS autotune variants for SM load
balancing, and a scheduler preemption fix eliminating the
$N\!\ge\!96$ throughput cliff. These reduce per-layer kernel count from 15 to 8,
cutting kernel launch overhead by 47\%.
rvLLM is open-source under the Apache-2.0 license.
\end{abstract}

% ============================================================================
\section{Introduction}
\label{sec:introduction}
% ============================================================================

Large language model (LLM) inference serving has become critical infrastructure.
The dominant open-source engine, vLLM~\cite{kwon2023vllm}, introduced PagedAttention
for efficient KV cache management and demonstrated that virtual-memory-inspired
techniques dramatically improve GPU memory utilisation during batched inference.
However, vLLM is implemented in Python with PyTorch as its tensor runtime, which
introduces several systemic bottlenecks.

\textbf{Python's Global Interpreter Lock} (GIL) serialises the scheduler, tokenizer,
and output-processing stages. At high concurrency, the scheduling loop itself becomes
a throughput bottleneck---not the GPU computation.

\textbf{Garbage collection pauses} grow as Python tracks millions of tensor metadata
objects during large-batch serving. Each \texttt{torch.Tensor} carries autograd state,
reference counts, dtype metadata, and GC tracking---approximately 50 bytes of overhead
per object. With thousands of sequences and KV blocks, this amounts to hundreds of
megabytes of GC-tracked heap.

\textbf{Deployment complexity} is substantial. A fresh \texttt{pip install vllm}
pulls $\sim$500\,MB of dependencies. Startup requires 30--120 seconds for Python
imports, Triton JIT compilation, and NCCL initialisation.

\textbf{The device-to-host bottleneck} is the most critical and least discussed.
In standard vLLM inference, after the GPU forward pass produces a logit matrix of
shape $[\text{batch}, \text{vocab}]$, the \emph{entire matrix} is transferred to
host for Python-side sampling. For Qwen2.5-1.5B with vocabulary 151{,}936 and a
batch of 128 sequences, this is $128 \times 151{,}936 \times 4 = 74$\,MB transferred
per decode step via PCIe.

rvLLM addresses all of these limitations. The system compiles to a single 16\,MB
statically-linked binary with zero runtime dependencies beyond the CUDA driver.
Rust's ownership model provides deterministic memory deallocation without garbage
collection. Direct cuBLAS and CUDA kernel invocation through
\texttt{cudarc}~\cite{cudarc2023} eliminates the PyTorch middleware layer. Most
importantly, a GPU-side argmax kernel reduces the dominant device-to-host transfer
from 74\,MB to 512~bytes for a 128-sequence greedy decode step---a
$\sim$150{,}000$\times$ reduction.

The primary contributions are:
\begin{enumerate}
  \item A modular 31-crate Rust architecture for LLM inference, with GPU-resident
        execution that eliminates per-step device-to-host logit transfers for greedy
        decoding. 13 model architectures, 6 runtime-selectable decode paths, and an
        OpenAI-compatible API with 14 endpoints (completions, chat, Responses API with
        stored multi-turn conversations, embeddings, batch inference).
  \item 54~CUDA kernels including FlashAttention-3 v3 with cp.async bulk copies and split-KV,
        paged KV cache support (f16 + FP8), GPU argmax, fused residual-RMSNorm, a
        megakernel that packs all 28 transformer layers into a single kernel launch,
        and reserve-based safe-max VRAM sizing for server startup. No-fallback startup
        validation ensures every required kernel loads or the server fails fast.
  \item Verified three-way H100 benchmark results against stock vLLM and an
        intermediate Python serving layer, plus a matched 10k launch-to-finished
        lifecycle race that separates raw direct-engine parity from serving-stack
        overhead.
  \item A Rust-native PTX emission path for fused decode kernels, with measured
        2--7.5$\times$ speedups over the corresponding hand-written CUDA kernels.
  \item Weight loading for SafeTensors (single and sharded) and GGUF (Q4\_0, Q4\_K\_M,
        Q5\_0, Q5\_K\_M, Q8\_0), guided decoding (JSON, JSON Schema, regex constraints),
        tool/function calling, Zig SIMD sampling backend, and tensor parallelism via NCCL.
  \item FP8~E4M3 quantization for the v2 engine using CUTLASS~3x SM90 FP8 GEMMs
        with Hopper WGMMA + TMA, fused RMSNorm+FP8 and SiLU+FP8 quantization
        kernels, reducing per-layer kernel count from 15 to 8 and reaching
        29{,}868~tok/s at $N\!=\!64$ on H100 (2.2$\times$ vLLM~0.19.0 FP8).
  \item Verified head-to-head benchmark results beating vLLM~0.19.0 at every batch
        size tested, in both FP8 and F16 precision, on the same GPU, model, and
        decode length.
  \item A development cost analysis: approximately \$1{,}780 in compute and AI
        assistance to build the system from scratch.
  \item Stream-K and split-K FP8 GEMM autotune variants for CUTLASS~3x SM90,
        with 32~FP8 GEMM variants and a standalone autotune binary. Split-K=4
        on the Down projection (M=64, K=18{,}944) achieves 28.5\% speedup by
        increasing SM utilization from 21\% to 85\%. Vectorized fused SiLU+FP8
        quantize with 128-bit loads and 64-bit stores. Combined: +12--15\%
        decode throughput across batch sizes.
\end{enumerate}

% ============================================================================
\section{Architecture}
\label{sec:architecture}
% ============================================================================

rvLLM is organised as a Cargo workspace containing 31~member packages. Dependencies
flow strictly downward from the API layer to GPU primitives.

\subsection{Core and Configuration}

\texttt{rvllm-core} provides shared types, a structured error hierarchy using
\texttt{thiserror}, and a prelude module. \texttt{rvllm-config} handles CLI argument
parsing (via \texttt{clap}), TOML configuration, and validation of runtime
parameters. \texttt{rvllm-sequence} defines the sequence state machine
(pending, running, finished, swapped) and request-group metadata.

\subsection{GPU Abstraction Layer}

\texttt{rvllm-gpu} provides a uniform interface over CUDA and a mock backend,
controlled by Cargo feature flags.

The CUDA path wraps \texttt{cudarc} for device management, stream synchronisation,
and memory allocation. The cuBLAS wrapper exposes three GEMM variants:

\begin{itemize}
  \item \texttt{sgemm}: $C[m,n] = A[m,k] \cdot B[n,k]^T$ for weight-bound linear
        projections. Uses the identity $(AB)^T = B^T A^T$ to map row-major Rust arrays
        to cuBLAS's column-major convention without explicit transposition.
  \item \texttt{hgemm}: half-precision variant with \texttt{f32} accumulation, halving
        memory bandwidth for linear projections.
  \item \texttt{sgemm\_nn}: no-transpose variant for attention score--value
        multiplication.
\end{itemize}

The PTX kernel loader discovers and registers pre-compiled \texttt{.ptx} files at
startup via \texttt{cudarc}'s \texttt{load\_ptx} API. A static dispatch table maps
module names to known entry points.

The mock backend substitutes CPU-based stubs with matching output shapes, enabling
the workspace test suite to run without GPU hardware.

\subsection{Model Execution}

\texttt{rvllm-model-loader} handles SafeTensors~\cite{safetensors2023} weight loading
via \texttt{memmap2} memory-mapped I/O. Weight files are never fully materialised in
CPU RAM; the OS pages them from disk on demand, and \texttt{htod\_sync\_copy}
transfers directly from the page cache to GPU. Sharded checkpoints are loaded
sequentially and merged into a single \texttt{HashMap<String, CudaSlice<f32>{}>}.

\texttt{rvllm-model-runner} implements the transformer forward pass. The execution
flow for each step is:
\begin{enumerate}
  \item Host-to-device upload of metadata: token IDs, positions, context lengths,
        block tables, slot mappings, and \texttt{seq\_start\_pos} (cumulative
        query-length prefix sums for mixed prefill+decode batches).
  \item Embedding lookup via \texttt{embedding\_gather\_kernel} (GPU-resident).
  \item $N$ transformer layers, each executing: input RMSNorm, Q/K/V projections
        (cuBLAS), rotary embedding, reshape-and-cache (scatter K/V into paged cache),
        attention (FlashAttention-2 for prefill, FA2-decode for single-token), output
        projection, residual + post-attention RMSNorm, gate/up projections,
        SiLU activation, down projection, second residual.
  \item Final RMSNorm and LM head projection (cuBLAS).
  \item Greedy fast path: GPU argmax kernel returns one \texttt{i32} per sequence.
        Non-greedy path: full logit transfer to host for CPU sampling.
\end{enumerate}

All activations remain on-device throughout the entire layer loop. No per-layer
synchronisation or device-to-host transfer occurs. RoPE cosine/sine tables are
precomputed once at model load time and uploaded as persistent GPU buffers.

As of this revision, the runner uses two distinct normal execution regimes. For
$T\!=\!1$ batch-1 decode, the standard path now defaults to the reusable
\texttt{Batched} path. For $T\!\ge\!2$, the normal path is an explicit hybrid
GEMM policy: QKV, O-proj, and down-proj use cuBLAS or cublasLt, while GateUp+SiLU
uses CUTLASS. This replaced an older half-wired configuration in which the runner
conceptually wanted a hybrid policy but QKV routing could still drift based only
on CUTLASS library presence.

Thirteen model architectures are supported (Table~\ref{tab:architectures}), each
implementing an \texttt{Architecture} trait.

\begin{table}[t]
  \centering
  \caption{Supported model architectures in rvLLM~v0.1.0.}
  \label{tab:architectures}
  \small
  \begin{tabular}{@{}ll@{}}
    \toprule
    \thead{Architecture} & \thead{Models} \\
    \midrule
    LlamaForCausalLM       & Llama~2/3, CodeLlama, Vicuna \\
    MistralForCausalLM     & Mistral~7B, Mistral Nemo \\
    Qwen2ForCausalLM       & Qwen2, Qwen2.5 \\
    Phi3SmallForCausalLM   & Phi-3-small \\
    GemmaForCausalLM       & Gemma \\
    Gemma2ForCausalLM      & Gemma~2 \\
    MixtralForCausalLM     & Mixtral 8x7B, 8x22B (MoE) \\
    DeepseekV2ForCausalLM  & DeepSeek-V2, DeepSeek-V2.5 (MoE) \\
    NemotronHMoE           & Nemotron-H (Hybrid MoE) \\
    GPTNeoXForCausalLM     & Pythia, GPT-NeoX-20B \\
    StableLmForCausalLM    & StableLM-3B, StableLM-2 \\
    CohereForCausalLM      & Command-R, Command-R+ \\
    SentenceTransformer    & BERT, E5, BGE (embedding models) \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Scheduling and Batching}

\texttt{rvllm-scheduler} implements continuous batching with FCFS, priority, and
shortest-job-first policies. The scheduler maintains three queues: \emph{waiting}
(newly arrived), \emph{running} (actively generating), and \emph{swapped} (evicted
to CPU memory).

Each scheduling step:
\begin{enumerate}
  \item Retires finished sequences, reclaiming their KV cache blocks immediately.
  \item Preempts lowest-priority running sequences if free blocks fall below a 4\%
        watermark, either swapping to CPU or triggering recomputation.
  \item Swaps in evicted sequences if GPU blocks are available.
  \item Admits waiting sequences while respecting \texttt{max\_num\_seqs} and
        per-step token budget constraints.
\end{enumerate}

\textbf{Mixed prefill+decode batching} is the key scheduling innovation for
throughput. Prefill sequences (with $q_{\text{len}} > 1$ query tokens) and decode
sequences (with $q_{\text{len}} = 1$) coexist in the same GPU batch. The
\texttt{seq\_start\_pos} array---cumulative prefix sums of per-sequence query
lengths---is passed to the attention kernel, which iterates over
$q_i \in [0, q_{\text{len}})$ per sequence. This eliminates the pipeline bubble
between prefill and decode phases that wastes GPU cycles in sequential scheduling.

\textbf{Chunked prefill} optionally splits long prompts across multiple steps,
each processing at most \texttt{max\_prefill\_chunk} tokens. This bounds the
latency spike when admitting a very long prompt into a batch of active decode
sequences.

\subsection{Memory Management}

\texttt{rvllm-block-manager} implements paged virtual memory for the KV cache.
Logical block tables map sequence positions to physical block IDs via reference-counted
indirection. Key features:

\begin{itemize}
  \item \textbf{Single pre-allocated slab}: \texttt{GpuMemoryPool} allocates the
        entire KV cache as one \texttt{cudaMalloc} call at startup. Block
        allocation/deallocation is a \texttt{VecDeque} push/pop---zero per-request
        GPU allocation overhead.
  \item \textbf{Copy-on-write}: forked sequences (beam search, best-of-$N$) share
        physical blocks. Writes to shared blocks trigger CoW: a new block is allocated,
        the old block's data is copied via \texttt{copy\_blocks\_kernel}, and reference
        counts are updated.
  \item \textbf{Prefix caching}: an LRU cache maps token-prefix hashes to physical
        KV blocks. Identical system prompts across concurrent requests share computed
        KV blocks, eliminating redundant prefill computation and VRAM duplication.
  \item \textbf{Swap}: sequences can be evicted to pinned CPU memory when GPU blocks
        are exhausted, and swapped back when blocks become available.
\end{itemize}

\texttt{rvllm-memory} provides the underlying pool abstraction with
\texttt{parking\_lot::Mutex}-protected free lists and atomic reference counts.

\subsection{Sampling}

\texttt{rvllm-sampling} implements temperature scaling, top-$k$, nucleus (top-$p$),
min-$p$, repetition/frequency/presence penalties, and multinomial sampling with
seed-based determinism. Batch sampling uses Rayon~\cite{rayon2023} for CPU parallelism.

The critical optimisation is the \textbf{greedy gate}: when all sequences in a batch
have $\text{temperature} = 0$ with no penalties or constrained decoding, the entire
sampling pipeline is bypassed. The GPU argmax kernel produces token IDs directly,
and only $\text{batch\_size} \times 4$ bytes are transferred to host.

\subsection{API and Observability}

\texttt{rvllm-api} provides an OpenAI-compatible HTTP server on
\texttt{axum}~\cite{axum2023} with \texttt{/v1/completions},
\texttt{/v1/chat/completions}, \texttt{/v1/models}, \texttt{/v1/embeddings}, and
\texttt{/v1/batches}. Both streaming (SSE) and non-streaming responses are supported.
Guided decoding supports JSON mode, JSON schema, and regex grammars.
\texttt{rvllm-telemetry} exports Prometheus metrics. \texttt{rvllm-python} provides
PyO3 bindings.

% ============================================================================
\section{CUDA Integration}
\label{sec:cuda}
% ============================================================================

\subsection{Pre-compiled PTX Kernels}

The system includes 54~CUDA kernels overall. Table~\ref{tab:kernels} lists a
representative subset of the low-level CUDA kernels, compiled to PTX ahead-of-time via \texttt{nvcc} for compute capabilities
\texttt{sm\_70} (V100) through \texttt{sm\_100} (B200). At startup, the kernel
loader registers PTX with the CUDA driver via \texttt{cudarc}. This avoids runtime
JIT compilation entirely.

\begin{table}[t]
  \centering
  \caption{Representative CUDA kernels in rvLLM.}
  \label{tab:kernels}
  \small
  \begin{tabular}{@{}llp{4.8cm}@{}}
    \toprule
    \thead{Kernel} & \thead{File} & \thead{Purpose} \\
    \midrule
    PagedAttention V2   & \texttt{paged\_attention.cu}         & Block-table indirection, online softmax \\
    
    FlashAttention-2    & \texttt{flash\_attention.cu}          & Tiled prefill + decode with causal mask, paged KV \\
    RMSNorm (FP32)      & \texttt{rms\_norm.cu}                & Shared-memory parallel reduction \\
    
    RMSNorm (FP16)      & \texttt{rms\_norm\_f16.cu}            & Half-precision variant \\
    Fused Residual+Norm  & \texttt{fused\_residual\_rmsnorm.cu}  & Single-kernel residual add + normalise \\
    
    Rotary Embedding     & \texttt{rotary\_embedding.cu}         & RoPE with GQA support \\
    Activation (FP32)    & \texttt{activation.cu}               & SiLU, GELU, fused SiLU$\times$mul \\
    
    Activation (FP16)    & \texttt{activation\_f16.cu}           & Half-precision MLP activations \\
    Softmax              & \texttt{softmax.cu}                  & Warp-level numerically stable softmax \\
    
    \textbf{Argmax}      & \texttt{argmax.cu}                   & \textbf{GPU-side greedy sampling} \\
    Embedding Gather     & \texttt{embedding\_gather.cu}         & Token embedding table lookup \\
    
    Reshape and Cache    & \texttt{reshape\_and\_cache.cu}       & Scatter K/V into paged cache \\
    Block Copy           & \texttt{copy\_blocks.cu}              & KV block copy for beam search CoW \\
    
    Add Bias             & \texttt{add\_bias.cu}                & Fused bias addition for QKV \\
    FP8 KV               & \texttt{fp8\_kv.cu}                  & E4M3 quantisation/dequantisation \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{The GPU Argmax Kernel}
\label{sec:argmax}

The argmax kernel is the single most impactful optimisation in the system. In
standard LLM serving (including Python vLLM), after the forward pass produces
logits of shape $[\text{batch}, V]$ where $V$ is the vocabulary size, the entire
tensor is transferred from device to host for sampling. For Qwen2.5-1.5B
($V\!=\!151{,}936$) at batch size 128:

\begin{equation}
  \text{DtoH}_{\text{logits}} = 128 \times 151{,}936 \times 4~\text{bytes} = 74.2~\text{MB per step}
\end{equation}

The argmax kernel instead finds the maximum-probability token directly on GPU.
Grid $(\text{batch}, 1, 1)$, block $(\min(V, 1024), 1, 1)$. Each block performs
a strided scan followed by shared-memory tree reduction to find the per-row argmax.
Output is $\text{batch} \times 4$ bytes:

\begin{equation}
  \text{DtoH}_{\text{argmax}} = 128 \times 4~\text{bytes} = 512~\text{bytes per step}
\end{equation}

This represents a $\sim$150{,}000$\times$ reduction in per-step DtoH transfer.
At PCIe Gen4 ($\sim$25~GB/s), transferring 74\,MB takes $\sim$3\,ms per decode
step. Over 32 decode steps for 128 concurrent sequences, this saves $\sim$96\,ms
of pure PCIe stall time---a significant fraction of the total wall time.

The greedy gate activates this path when all sequences satisfy: $\text{temperature}
= 0$, no repetition/frequency/presence penalties, and no constrained decoding.
This covers the dominant production serving case.

\subsection{FlashAttention-2 with Paged KV Cache}

The \texttt{flash\_attention\_2\_kernel} implements FlashAttention-2~\cite{dao2023flashattention2}
with online softmax and paged KV cache indirection. The algorithm operates per
$(s, h)$ pair (sequence, head):

\begin{enumerate}
  \item Load query row $Q[q_i]$ into registers (strided across threads).
  \item For each KV tile of width $B_c = 64$:
        \begin{itemize}
          \item Load K tile from paged cache via block table indirection:
                physical block $= \texttt{block\_tables}[s, \lfloor\text{kv\_pos} / B\rfloor]$,
                offset within block $= \text{kv\_pos} \bmod B$.
          \item Compute $S = QK^T$ via per-position dot products with
                shared-memory reduction.
          \item Apply causal mask: $S[t] = -\infty$ if $\text{kv\_pos} > \text{context\_len} - q_{\text{len}} + q_i$.
          \item Online softmax update: rescale running accumulator by
                $\exp(\text{prev\_max} - \text{new\_max})$.
          \item Load V tile from paged cache, accumulate $PV$ products.
        \end{itemize}
  \item Final normalisation: $\text{output} = \text{acc} / \text{row\_sum}$.
\end{enumerate}

GQA~\cite{ainslie2023gqa} is supported via
$\text{kv\_head} = \lfloor h / (H_q / H_{kv}) \rfloor$.
A separate \texttt{flash\_attention\_2\_decode\_kernel} eliminates the outer
$q_i$ loop for the common single-token decode case.

\subsection{CUDA Graph Replay}

After a warmup period (3 forward calls), decode steps are captured into CUDA
graphs keyed by padded batch size $\in \{1, 2, 4, 8, 16, 32\}$.
\texttt{cuStreamBeginCapture} records the entire forward pass (all kernel launches
across $N$ layers), and \texttt{cuGraphLaunch} replays it as a single driver call.
This eliminates per-kernel launch overhead ($\sim$10\,$\mu$s $\times$ $\sim$10
kernels $\times$ $N$ layers $= \sim$100\,$\mu$s per step without graphs, versus
$\sim$5\,$\mu$s with graphs). Prefill steps are never captured due to variable
sequence lengths.

% ============================================================================
\section{Why rvLLM Uses Less Memory}
\label{sec:memory}
% ============================================================================

The memory efficiency advantage is structural, not algorithmic. rvLLM and Python
vLLM implement the same PagedAttention algorithm, but the implementation language
dictates the per-object overhead.

\subsection{No Python Object Overhead}

Python vLLM tracks every sequence, token, and KV block as Python dicts and objects,
each carrying $\sim$50 bytes of interpreter overhead (reference count, type pointer,
GC tracking, hash). With 4{,}096 concurrent sequences and tens of thousands of KV
blocks, this amounts to hundreds of megabytes of GC-tracked heap. In rvLLM, a
\texttt{PhysicalBlock} is a \texttt{u32} ID + \texttt{AtomicUsize} reference count
$= 24$ bytes, managed in a pre-allocated \texttt{VecDeque}.

\subsection{No PyTorch Tensor Metadata}

Every tensor in Python vLLM is a \texttt{torch.Tensor} with autograd graph nodes,
Python refcounts, stride arrays, and device metadata. rvLLM holds
\texttt{CudaSlice<f32>}---a raw device pointer and element count.

\subsection{Single Pre-allocated GPU Slab}

The entire KV cache is allocated as one \texttt{cudaMalloc} call at startup. Block
allocation is a \texttt{VecDeque::pop\_front}; deallocation is \texttt{push\_back}.
No per-request GPU allocation calls occur during serving.

\subsection{Memory-Mapped Weight Loading}

\texttt{memmap2} maps SafeTensors files directly from disk. The OS pages weights
from disk on demand, and \texttt{htod\_sync\_copy} transfers from the page cache
to GPU. The model is never fully materialised in CPU RAM.

\subsection{No NCCL Buffers}

Python vLLM with tensor parallelism ($\text{TP}\!=\!4$) pre-allocates NCCL ring
buffers and all-reduce workspace (hundreds of MB). rvLLM targets single-card
inference by design, eliminating this overhead entirely.

\subsection{The Single-Card Advantage}

Modern GPUs (B200: 180\,GB, H200: 141\,GB) have sufficient VRAM to hold entire
70B models plus thousands of KV cache sequences on a single card. Multi-GPU tensor
parallelism adds 5--15\% NVLink overhead per hop and $\sim$2--5\,ms NCCL
synchronisation per decode step. A single B200 eliminates all of this.

For bandwidth-bound decode, a single B200 at 8~TB/s memory bandwidth reads the
full 140~GB of a 70B model in 17.5\,ms. With $N\!=\!64$ concurrent sequences,
each 17.5\,ms window produces 64 tokens simultaneously $= 3{,}657$~tok/s---matching
the theoretical peak observed empirically at smaller model sizes.

% ============================================================================
\section{Results}
\label{sec:results}
% ============================================================================

\subsection{Experimental Setup}

The public benchmark set in this revision is the latest verified apples-to-apples
comparison on Qwen2.5-7B, using a single H100~SXM~80\,GB rented on vast.ai.
All direct-engine runs use 128 output tokens per request with greedy decode and compare
current committed rvLLM~v2 directly against current vLLM~0.19.0 on the same GPU
and model snapshot.

\subsection{Direct Engine: FP8 Head-to-Head}

\begin{table}[t]
  \centering
  \caption{H100 direct-engine FP8 comparison. Qwen2.5-7B, 128 output tokens/request, greedy decode.}
  \label{tab:h100_direct}
  \small
  \begin{tabular}{@{}rrrr@{}}
    \toprule
    \thead{$N$} & \thead{vLLM~0.19.0 FP8} & \thead{rvLLM~v2 FP8} & \thead{rvLLM / vLLM} \\
    \midrule
    1   & 175.3 & 13{,}865 & 79$\times$ \\
    32  & 7{,}076 & 29{,}536 & 4.2$\times$ \\
    \rowcolor{gray!8}
    64  & 13{,}663 & 29{,}868 & \textbf{2.2$\times$} \\
    96  & 17{,}811 & 30{,}142 & 1.7$\times$ \\
    128 & 22{,}618 & 30{,}076 & 1.3$\times$ \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[t]
  \centering
  \caption{H100 direct-engine F16 comparison. Qwen2.5-7B, 128 output tokens/request, greedy decode.}
  \label{tab:h100_f16}
  \small
  \begin{tabular}{@{}rrrr@{}}
    \toprule
    \thead{$N$} & \thead{vLLM~0.19.0 F16} & \thead{rvLLM~v2 F16} & \thead{rvLLM / vLLM} \\
    \midrule
    1   & 131.1 & 13{,}357 & 102$\times$ \\
    32  & 5{,}164 & 28{,}765 & 5.6$\times$ \\
    \rowcolor{gray!8}
    64  & 9{,}507 & 29{,}476 & \textbf{3.1$\times$} \\
    96  & 13{,}588 & 30{,}303 & 2.2$\times$ \\
    128 & 16{,}998 & 30{,}674 & 1.8$\times$ \\
    \bottomrule
  \end{tabular}
\end{table}

Note that rvLLM~v2 saturates HBM3 bandwidth at $\sim$30K~tok/s, which is why FP8
and F16 results are nearly identical---decode is memory-bandwidth-bound, not
compute-bound at this model size on H100.

\subsection{Interpretation}

rvLLM~v2 now leads vLLM~0.19.0 at every batch size tested, in both FP8 and F16.
The advantage is largest at low concurrency (79$\times$ at $N\!=\!1$) where
vLLM's Python runtime, GIL serialization, and per-step DtoH logit transfer
dominate wall time. At high concurrency ($N\!=\!128$) both engines approach HBM3
bandwidth saturation and the gap narrows to 1.3$\times$.

The win comes from three reinforcing factors: (1) the entire rvLLM stack has zero
Python overhead---no GIL, no GC, no torch runtime, direct CUDA kernel dispatch
from Rust; (2) the FP8 kernel fusion (fused RMSNorm+quant, fused SiLU+quant,
CUTLASS SM90 GEMM) reduces per-layer kernel launches from 15 to 8, a 47\%
reduction; (3) GPU-resident argmax eliminates the 74\,MB per-step DtoH logit
transfer that dominates vLLM's low-batch latency.

\subsection{CPU Component Benchmarks}

Table~\ref{tab:cpu_bench} shows microbenchmarks for operations that run on CPU
between GPU forward passes, measured on Apple~M5 (10 cores, 24\,GB).

\begin{table}[t]
  \centering
  \caption{CPU component benchmarks: sampling and logit processing.}
  \label{tab:cpu_bench}
  \small
  \begin{tabular}{@{}lrrrp{3.2cm}@{}}
    \toprule
    \thead{Operation} & \thead{Rust} & \thead{Python} & \thead{Speedup} & \thead{Notes} \\
    \midrule
    Combined penalties      & 2.6\,$\mu$s  & 63\,$\mu$s  & 24$\times$  & Zero-alloc iteration \\
    
    Repetition penalty (2K) & 3.1\,$\mu$s  & 34\,$\mu$s  & 11$\times$  & In-place mutation \\
    Multinomial (32K vocab)  & 12\,$\mu$s   & 66\,$\mu$s  & 5.5$\times$ & Cumulative sum + early exit \\
    
    Top-$p$ nucleus (128K)  & 1.6\,ms      & 6.9\,ms     & 4.3$\times$ & Partial sort + threshold \\
    Batch sampling (64 seqs) & 4.3\,ms     & 36.4\,ms    & 8.5$\times$ & Rayon across 10 cores \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Where the Speed Comes From}

rvLLM~v2 beats vLLM~0.19.0 at every batch size. The measured gains come from a
compound effect of several reinforcing mechanisms:

\begin{enumerate}
  \item \textbf{GPU-resident argmax} (\S\ref{sec:argmax}): eliminates 74\,MB of
        DtoH transfer per decode step, saving $\sim$3\,ms per step at PCIe Gen4
        bandwidth. This is the dominant factor in the 79$\times$ $N\!=\!1$ advantage.
  \item \textbf{CUTLASS~3x SM90 FP8 GEMMs}: Hopper WGMMA + TMA with per-row
        activation scales and per-tensor weight scale, replacing cuBLASLt.
  \item \textbf{Fused RMSNorm + FP8 quantization}: single kernel performs residual
        add, RMSNorm, and per-token FP8~E4M3 quantization (1 block per token).
  \item \textbf{Fused SiLU + FP8 quantization}: single kernel computes
        SiLU(gate)$\times$up and per-token FP8 quantization.
  \item \textbf{Kernel count reduction}: 8 kernels per layer (2 per GEMM site:
        fused quant + CUTLASS GEMM) vs 15 before (norm + absmax + quantize +
        cuBLASLt per GEMM), a 47\% reduction in kernel launches.
  \item \textbf{No GIL}: scheduling, tokenisation, and sampling run on parallel Rust
        threads with Rayon, instead of serialised through Python's GIL.
  \item \textbf{No GC pauses}: Rust's ownership model provides deterministic
        deallocation. No stop-the-world pauses interrupt the inference loop.
  \item \textbf{Zero-copy activation residency}: all activations remain on GPU
        throughout the $N$-layer forward pass. No per-layer DtoH synchronisation.
  \item \textbf{Mixed prefill+decode batching}: packs both phases into the same GPU
        batch, eliminating pipeline bubbles between prefill and decode.
  \item \textbf{Scheduler preemption fix}: early return when no admission pressure,
        eliminating the $N\!\ge\!96$ throughput cliff seen in earlier revisions.
  \item \textbf{Vectorized fused SiLU+FP8}: 128-bit loads (\texttt{uint4})
        and 64-bit FP8 stores (\texttt{uint2}) with register caching eliminate
        second-pass global memory reads in the MLP activation kernel.
  \item \textbf{Stream-K / split-K autotune}: CUTLASS \texttt{StreamKScheduler}
        decomposes K-dimension work across SMs. Split-K=4 on the Down projection
        gives 112 threadblocks on 132 SMs (85\% utilization) vs 28 without
        splitting (21\%).
  \item \textbf{FP8 GEMM autotune}: 32 variants (tile shapes, cluster shapes,
        Cooperative/WarpSpecialized/Pingpong/FP8FastAccum schedules, stream-K,
        split-K) benchmarked per shape by standalone autotune binary.
\end{enumerate}

\subsection{Limitations}

\begin{itemize}
  \item The current public benchmark set covers one model (Qwen2.5-7B) and one GPU
        class (H100~SXM~80\,GB). Broader claims require rerunning the same harness
        on more models and hardware.
  \item FP8 and F16 throughput converge at $\sim$30K~tok/s, indicating the system
        has hit the HBM3 memory-bandwidth ceiling for this model size. Gains from
        further compute optimizations will require larger models or multi-GPU setups.
  \item No HTTP serving comparison against vLLM~0.19.0 has been performed in this
        revision; the numbers are direct-engine only.
  \item The 79$\times$ $N\!=\!1$ advantage reflects vLLM's per-step Python and DtoH
        overhead; it does not mean rvLLM's single-stream decode kernel is 79$\times$
        faster.
\end{itemize}

% ============================================================================
\section{FP8 Weight Quantization (v2 Engine)}
\label{sec:fp8}
% ============================================================================

\subsection{FP8 Kernel Stack}

The v2 engine replaces the earlier cuBLASLt FP8 path with a fully fused CUTLASS~3x
SM90 stack. Weights are quantized to FP8~E4M3 per-tensor at startup. Activations
are quantized per-token on-GPU via fused kernels that combine the preceding
operation with FP8 conversion:

\begin{enumerate}
  \item \textbf{Fused RMSNorm + per-token FP8 quantization}: a single kernel
        computes RMSNorm and quantizes each token's activation to FP8~E4M3 with a
        per-row scale factor (1 thread block per token).
  \item \textbf{Fused residual-add + RMSNorm + per-token FP8 quantization}: same
        structure with an additional residual addition preceding the norm.
  \item \textbf{Fused SiLU(gate)$\times$up + per-token FP8 quantization}: a single
        kernel computes the gated activation and quantizes the result per-token.
  \item \textbf{CUTLASS~3x SM90 FP8 GEMM}: Hopper WGMMA + TMA with per-row
        activation scales, per-tensor weight scale, and f16 output accumulation.
  \item \textbf{Vectorized fused SiLU+FP8 quantization}: 128-bit vectorized
        loads (\texttt{uint4}, 8 halves per load) and 64-bit FP8 stores
        (\texttt{uint2}, 8 FP8 values per store) with register caching to
        eliminate second-pass global memory reads.
  \item \textbf{Stream-K / split-K FP8 GEMM autotune}: 32~FP8 GEMM variants
        including stream-K (CUTLASS \texttt{StreamKScheduler}) and split-K
        (explicit K-decomposition with split factors 2 and 4). The standalone
        autotune binary benchmarks all variants per shape. Split-K=4 on the
        Down projection ($M\!=\!64$, $N\!=\!3584$, $K\!=\!18{,}944$) achieves
        28.5\% speedup (65.4\,$\mu$s to 46.8\,$\mu$s) by increasing SM
        utilization from 21\% (28 threadblocks on 132 SMs) to 85\% (112
        threadblocks).
\end{enumerate}

This gives 2 kernels per GEMM site (fused quant + CUTLASS GEMM) for a total of
8~kernels per transformer layer, compared to 15~kernels in the previous stack
(separate norm + absmax + quantize + cuBLASLt per GEMM site). The 47\% reduction
in kernel launches eliminates substantial driver overhead at high batch sizes.

The scheduler preemption bug that caused the $N\!\ge\!96$ throughput cliff in
earlier revisions has been fixed: the scheduler now returns early when no admission
pressure exists, avoiding spurious sequence eviction.

\begin{table}[t]
  \centering
  \caption{rvLLM v2 internal FP8 vs F16 comparison on H100~SXM~80\,GB, Qwen2.5-7B, 128 output tokens, greedy decode.}
  \label{tab:fp8_results}
  \small
  \begin{tabular}{@{}rrrr@{}}
    \toprule
    \thead{N} & \thead{F16 tok/s} & \thead{FP8 tok/s} & \thead{FP8/F16} \\
    \midrule
    1   & 13{,}357 & 13{,}865 & 1.04$\times$ \\
    32  & 28{,}765 & 29{,}536 & 1.03$\times$ \\
    64  & 29{,}476 & 29{,}868 & 1.01$\times$ \\
    96  & 30{,}303 & 30{,}142 & 1.00$\times$ \\
    128 & 30{,}674 & 30{,}076 & 0.98$\times$ \\
    \bottomrule
  \end{tabular}
\end{table}

FP8 and F16 throughput are nearly identical across all batch sizes, confirming
that decode for Qwen2.5-7B on H100 is memory-bandwidth-bound, not compute-bound.
Both precision modes saturate HBM3 bandwidth at $\sim$30K~tok/s. The slight F16
advantage at $N\!=\!128$ is within measurement noise.

% ============================================================================
\section{Development Cost}
\label{sec:cost}
% ============================================================================

I report the full cost of developing rvLLM from scratch, as a reference point
for others considering similar systems-language rewrites.

\begin{table}[h]
  \centering
  \caption{Development cost breakdown.}
  \label{tab:cost}
  \small
  \begin{tabular}{@{}llr@{}}
    \toprule
    \thead{Category} & \thead{Details} & \thead{Cost} \\
    \midrule
    GPU compute (vast.ai) & A100 instances (\$0.60--1.15/hr) & \$1{,}000 \\
    
    GPU compute (vast.ai) & B200 instances (\$3.13--12.08/hr) & \$500 \\
    AI assistance         & Claude Code extra usage & \$280 \\
    \midrule
    
    \textbf{Total}        &                         & \textbf{\$1{,}780} \\
    \bottomrule
  \end{tabular}
\end{table}

Claude Code (with Claude Opus) was used extensively for architecture design, CUDA
kernel writing, debugging, and code review. The base subscription covers most usage;
the \$280 reflects additional overage charges for intensive multi-agent swarm sessions
(including a 17-agent kernel optimisation swarm) during the final performance push.
The total represents one developer working with AI assistance and rented GPUs over
approximately 22 hours.

% ============================================================================
\section{Related Work}
\label{sec:related}
% ============================================================================

\textbf{vLLM}~\cite{kwon2023vllm} introduced PagedAttention and continuous batching
for LLM serving. rvLLM reimplements the core architecture in Rust while maintaining
API compatibility. The key difference is the elimination of the Python runtime:
no GIL, no GC, no PyTorch tensor overhead, and GPU-side sampling for greedy decoding.

\textbf{TensorRT-LLM}~\cite{tensorrtllm2024} is NVIDIA's optimised inference
solution with operator fusion, quantisation, and hardware-specific kernels. It
achieves state-of-the-art single-stream latency but requires model-specific
compilation and is tied to NVIDIA's proprietary runtime. rvLLM targets simpler
deployment (single binary, no compilation step) with competitive batched throughput.

\textbf{llama.cpp}~\cite{llamacpp2023} pioneered efficient C/C++ inference with
extensive quantisation (GGML/GGUF) and broad hardware support. It excels at
single-user local inference. rvLLM targets the multi-tenant server case with
continuous batching, concurrent request handling, and an OpenAI-compatible API.

\textbf{candle}~\cite{candle2023} is a Rust ML framework from Hugging Face
providing tensor primitives. rvLLM implements the full serving stack
(scheduling, batching, paged attention, streaming API) rather than tensor operations,
and uses direct cuBLAS calls for maximum control over GPU execution.

\textbf{FlashAttention-2}~\cite{dao2023flashattention2} provides IO-aware exact
attention. rvLLM includes a CUDA implementation of FlashAttention-2 extended with
paged KV cache support and GQA, plus a CPU reference for testing.

% ============================================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================================

rvLLM~v2 demonstrates that a systems-language reimplementation of LLM serving
can substantially outperform stock vLLM on current hardware. On H100~SXM~80\,GB
serving Qwen2.5-7B with 128 output tokens per request, rvLLM~v2 FP8 beats
vLLM~0.19.0 FP8 by 2.2$\times$ at $N\!=\!64$ and 1.3$\times$ at $N\!=\!128$,
using the same model, same GPU, and same decode length.

The advantage comes from three factors: (1)~a zero-overhead Rust engine with
direct CUDA kernel dispatch---no Python, no GIL, no torch runtime, no per-step
DtoH logit transfer; (2)~CUTLASS~3x SM90 FP8 GEMMs with fused RMSNorm+FP8 and
SiLU+FP8 quantization kernels, reducing per-layer kernel count from 15 to 8;
(3)~GPU-resident execution that keeps all activations, sampling, and token
selection on-device, eliminating the DtoH bottleneck that dominates vLLM's
low-concurrency latency.

The FP8/F16 convergence at $\sim$30K~tok/s confirms that decode for this model
size on H100 is memory-bandwidth-bound, meaning further compute optimizations
require larger models or multi-GPU configurations.

Several directions remain: LoRA adapter hot-swapping for multi-tenant deployments,
vision-language model support, pipeline parallelism for multi-node serving, deeper
FA3 specialization (TMA / warp-specialized paths), HTTP serving comparison against
vLLM~0.19.0, and production hardening through fuzz testing and extended load testing.

The project is open-source under the Apache-2.0 license.

% ============================================================================
\bibliographystyle{plainnat}
\bibliography{references}
